---
title: "KEmodel"
author: "Alan Fortuny Sicart"
output:
  html_document:
    df_print: paged
  pdf_document: default
  word_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
## Introduction

There are serious concerns regarding the likelihood of a global collapse that have multiple origins.

From one side, we are still relying heavily on fossil fuels to feed our economy and mantain our societal complexity,
and its gettings harder and harder to keep up with the supply of oil, coal and gas. In order words, the time of abundant
and cheap energy is likely to end.

In parallel to this, humand caused global warming is going to make climate more extreme, increasing the amount of droughts and floods, and hence agriculture yields and basic infraestructure are under risk in many areas in the world.

A growing population, that aims to replicate the consumption patters from western societies, will find that the ladder to great consumption does not correlate to societal cohesion or even hapinness, as we see already. Frustation is guaranted, as the Earth cannot support the level of population that are currently or in the future if dramatic reductions on consumption and energy use are taken, specially in those areas when a significant amount of material output is in place.

The loss of biodiversity is not only a loss to our natural treasures and other species, but also a serious risk to our health, in the form of pandemics, polluted air or water, and a general loss of our resilience.
The growing challenges and accumulated damage on the soil, water and other living organisms are not resulting is the elimination of poverty, the moderation of inequality and the reduction in the amount of stress and anxiety.

Growth as a policy to fix the previously mention issues is very flawed, as it is resource innefective, given the growing concentration of the benefits of that growth, and even the costs of that growth. And price systems cannot fix the scale problem we have at hand. There has to be limits in the amount of natural capital destruction, and the amount of inequality accepted if we aim to have a resilient society in terms of resource capacity but also in terms of social cohesion. Collapse could come in the form of wars due to the extreme stress of some parts of society, or in the form of lack of resource available to coppe with our growing and apparently insatiable needs.

The Steady State Economics framework focus on the qualitative and not only quantitative aspect of development. Recognizing that not all growth is good, and that in a limited planet governened by the laws of thermodynamics, consumption and population must stagnated, and development have not only material dimensions but also qualitative ones as reflected in health, culture, social cohesion... While innovation is much needed to be more efficient and improve our lives, there must be a definition of enough, to avoid collapse and suffering it is important to shape our policy to limit costly growth, and give the chance to countries still in development face to take the path of enough, and not chasing the infinite race of more is better.

A economy and society that is more reslient, just and even happier is possible, as no growth does not mean no development. An economy rich in time, health, socieal relationships, fredoom, justice and full of life and biodiversity is worth exploring alternative paradigms to growth, even if that requires more complex and out of the box tools.

As economists we will asses the tradeoffs that exist between growth, consumption, inequality, natural capital...but taken a long term view, and not only maximizing a very short time window. Our intuition tell us that a better path for development is possible, where natural capital depletion is stopped, inequality is moderated, and moderate amounts of inequality are achieved. 

The following models: The Predator Prey Model, The Handy Model and the Capital and Environment model, give us the intuitions of how
certain actions on inequality, consumption, taxation and regulation of natural capital extraction, are not only necessary to avoid collapse but also Pareto efficient. 

In the current paper, no empirical estimation will take place, only though games based on parameters given by the literature, but in the next research more empirical testing on the recommendations and results provided in that document will be under examination.


## Predator Prey model (Lotka 1925, Volterra 1926)

To start and get familiar with the modeling approaches, the predator prey model is a simple yet valuable tool.

Suppose we have two species, first the human as a predator, and nature as the prey. 

The dynamics between those can be explain by the following equations:

$$ 
\dot{x} = (ay)x-bx  \   [1]
$$

$$ 
\dot{y} = (cy)-(dx)y  \   [1]
$$
Note that the change in the amount of humans x, is a function of the amoun of natural stock,y, times some constant
a which is the birth rate of x minus b which is the death rate.

By the same tolken, nature regenerates at rate c, but is reduced at rate d, times the amount of humans x.

Rather than having an stable equilibrium, there is a series of variations around the equilibrium values:

$$
x_{e} = c/b   \   [2]
$$
$$
y_{e} = b/a \   [2]
$$

## Ploting the system behaviour

In the following, we will see the behaviour of the system for the following values:


```{r}

a = 3*10^-5
b = 2*10^−2
c = 3*10^-2
d = 2*10^-4
xo=10^2
yo=10^3
T = 0:1000
xe = c/b
ye = b/a

simulate <- function(d=2*10^-4,c=3*10^-2) {
#Empty matrix with the evolution of the variables

simulation = as.data.frame(matrix(ncol=5,nrow=length(T)))
colnames(simulation) <- c("Time","humand_eq","nature_eq","human","nature")
simulation$Time <- T
simulation$humand_eq <- xe
simulation$nature_eq <- ye

# Running a loop for T

for (t in T){
  
  if(t==0){
    y=yo
    x=xo
    xdot = a*yo*xo-b*xo
    ydot = c*yo-d*xo*yo
  }else{
    xdot = a*y*x-b*x
    ydot = c*y-d*x*y
    y = y+ydot
    x = x+xdot
    simulation[t,c("human")] <- y
    simulation[t,c("nature")] <- x
  }
}
return(simulation) 
 
}

simulation <-simulate()

```


We can now plot the resuts based on the simulation

```{r}
matplot(simulation[,4:5], type = c("b"),pch=1,xlab = c("year"), ylab = c("population/stock"),
main ="Human and nature Scenario baseline") #plot
legend("topright", legend = colnames(simulation)[4:5], col=2:1, pch=1)

```


This apparently simple model show interesting series of abundancy and almost collapse, and its figure, clearly depend upon:

   1) The **relative size of the natural capital**, which is y/x
   2) The **consumption need per humand**, defined by d
   3) The **rate of regeneration of nature**, defined by c
   

Let's draw some scenarios:

   A) The consumption per human is 10 times greater
  

```{r}

simulation <-simulate(d=d*10)
matplot(simulation[,4:5], type = c("b"),pch=1,xlab = c("year"), ylab = c("population/stock"),
main ="Human and nature Scenario A high consumption") #plot
legend("topright", legend = colnames(simulation)[4:5], col=2:1, pch=1)


```

   
    B) The rate of regeneration of nature is 5 times faster than baseline

```{r}

simulation <-simulate(c=c*5)
matplot(simulation[,4:5], type = c("b"),pch=1,xlab = c("year"), ylab = c("population/stock"),
main ="Human and nature Scenario B fast recovery") #plot
legend("topright", legend = colnames(simulation)[4:5], col=2:1, pch=1)


```


In the baseline scenario, a series of growth and collapse periods are happening, on average every 250 years. If consumption is accelerated, population is lower and breakdown is almost permanent, while when recovery is accelerated, given the fact that population grows at a proportion of the stock of nature, collapses still happen although less often thatn in the previous scenarios.

What can we do to avoid collapse?

If we stablish a permanent amount of population, after a threshold, for example at xe, we should be able to avoid collapse. To to that we will set:

When 

$$ 
 y < x*d \  [3]
$$
Then:

$$ 
\dot{x} = 1-y/(x*d)   \  [3]
$$

Otherwise

$$ 
 \dot{x} =0 \  [3]
$$

Let's create another formula with fix population stock, when natural capital is enough:



```{r}

a = 3*10^-5
b = 2*10^−2
c = 1*10^-2
d = 1*10^-4
xo=10^2
yo=10^3
T = 0:1000

simulate_fix_consumption <- function(d=2*10^-4,c=3*10^-2) {
  #Empty matrix with the evolution of the variables
  
   simulation = as.data.frame(matrix(ncol=5,nrow=length(T)))
  colnames(simulation) <- c("Time","humand_eq","nature_eq","human","nature")
  simulation$Time <- T
  simulation$humand_eq <- xe
  simulation$nature_eq <- ye
  
  # Running a loop for T
  
  for (t in T){
    
    if(t==0){
      y=yo
      x=xo
      
    if(y<x*d*yo){
      
      xdot=-x*(y/(x*d*yo))
      
    }else{
      if(x<xo){
        xdot= min(a*x,xo-x)
        
      }else{
        xdot =0
      }
     
    }  
     
      ydot = c*yo-d*xo*yo
    }else{
      if(y<x*d*yo){
        
        xdot=-x*(y/(x*d*yo))
        
      }else{
        
        xdot =0
      }  
      ydot = c*y-d*x*yo
      y = y+ydot
      x = x+xdot
      simulation[t,c("human")] <- x
      simulation[t,c("nature")] <- y
    }
  }
  return(simulation) 
  }

simulation <-simulate_fix_consumption (c = 1*10^-2,d = 1*10^-4)

```




```{r}
matplot(simulation[,4:5], type = c("b"),pch=1,xlab = c("year"), ylab = c("population/stock"),
main ="Human and nature Scenario fix population and consumption rate") #plot
legend("topright", legend = colnames(simulation)[4:5], col=2:1, pch=1)

```


If we fix consumption to be d for all citizens, and maximum population to be 100, for a regenartion rate of 1% anually and consumption per capita of 0.01 units of the natural capital, is a steady state, where xe=xo, ye=yo for 1000 years, and consumption is smoothed for the whole period.Small deviations of such values can be catrastophic for the environment or the humanity in the long run.


```{r}
simulation <-simulate_fix_consumption (c = 1*10^-2,d = 8^-4)
matplot(simulation[,4:5], type = c("b"),pch=1,xlab = c("year"), ylab = c("population/stock"),
main ="Human and nature Scenario fix population but unestable consumption rate") #plot
legend("topright", legend = colnames(simulation)[4:5], col=2:1, pch=1)
```

Is that model satysfying, well, it is certainly useful but also very limited.
In the baseline scenario, while collapse is avoided there is a lot of societal pain, as population is forced to go down as available natural capital is not enough for everyone. Total collapse did not happen, but at the price of high population drifts,
something no of us really want.

In the scenario of fix population and consumption, we do achieve equilibrium, but is highly unstable, any slight deviation from the optimal level of consumption but break down the system and increase dramatically the amount of deaths.

In any case, the inequality dimensions of the problem are not consider, as well as the possibility to save wealth when extraction is not longer possible (or a shock happens). That could first reduce total consumption, as the elites will see their consumption cap to a boundary, freeing resources for the poor, and also savings and accumulation can protect population for down times, hence natural cycles do not affect population numbers.

The Handy Model will include the inequality into the equation and wealth accumulation, so we can understand how this interact with the environment. 


## The Handy Model (Motesharrei, Rivas, Kalnay 2014)

The Handy Model, brings the possibility to accumulate wealth, that can be consumed later, called wealth.

Another much needed feature in such modeling is the fact that wealth and consumption is not evenly distributed, and two types of population are created, the Elites with a much greater amount of wealth and consumption, and the Commoners, with much less.
We see globally that a small percentage of the population concentrates most of the wealth and consumption of natural resources.

The model hence becomes more complex than the Predator Pray model, as two equations are needed to model predator population, one for the elites and one for the commoners, one more equation for the natural capital and last for the wealth accumulation.


$$ 
\dot{x_C} = \beta_Cx_C-\alpha_Cx_C\   [4]
$$

$$ 
\dot{x_E} = \beta_Ex_E-\alpha_Ex_E\   [5]
$$


$$ 
\dot{y} = \gamma  y (\lambda-y)-\delta x_C  y  \   [6]
$$

$$ 
\dot{w} = \delta x_C  y - C_C-C_E \   [7]
$$


**Model Description**

The population is divided by xe[5] and xc[4], being the population of elites and commoners.Beta and alpha represent the birth and death rate.

Beta can be assumed to be the same for both classes, but alpha depends on the available wealth.

Note that although natural capital takes multiple forms : materials, water, soil, energy... we aggregate into one single number.


The equation of nature [6] containts a regeneration term:


$$ 
 \gamma  y (\lambda-y)
$$
and a depletion term:

$$ 
\delta x_C  y  \   
$$

Note that maximum regeneration takes place when:

$$ 
\ {y} = \delta/2 \ 
$$
The depletion side contains the pollution involved in production.The depletion is proportional to the number of workers, without any reference to technological changes or the impact of physical capital. We will see in the next model what are the implications of not adding endogenous technology and capital into the depletion term.In this model, only commoners produce.

The Handy model omits the impact on technological change as according to their authors "effects cancel out". The KE model presented later show the impact of having engoneous technological progress and it argues agains the destinty that improvements in efficiency will inevitably go to increases in consumption as the Handy model suggest. History is not destiny, and technology can play a significant role, even if it is far from what the infinit growth defenders try to make us believe.


If we look at the wealth accumulation equation [7], wealth increases with production at rate 


$$ 
\delta x_C  y 
$$

and decreases with total consumption

$$ 
C_C+C_E \   
$$

The consumption of commoners is equal to:

$$ 
C_C= min(1,w/w_f)sx_c \ [8]   
$$

Where s is the subsistence salary, and wf is the minimum value of wealth that if below, famine start


The consumption of elites is equal to:

$$ 
C_E= min(1,w/w_f)\kappa sx_C \ [9]   
$$

Where kappa is an exogenous value that defines the times the elites gt from the minimum salary s. As recognize by the authors this is a limitation of the model as it does not explain the sources of inequality and the chances to change over time. for the time being is a simple way to model the impact of some
persistent inequality into our environment dynamics.

The minimum wealth to avoid famine is expressed as:

$$ 
w_f = \rho x_C +\kappa \rho x_E   \   [10]   
$$

Note this is defined by rho, the minimum consumption for survival, where the elites got kappa times of it per person, in this model.

The death rates are determined dynamically as follows:


$$ 
\alpha_C = \alpha_m + max(0,1-C_c/sx_C)(\alpha_M - \alpha_m)  \   [11]   
$$

$$ 
\alpha_E = \alpha_m + max(0,1-C_E/sx_E)(\alpha_M - \alpha_m)  \   [12]   
$$

A healthy value of death rate (natural) has the subscript m, while M happens when a famine starts. Death rate can be also interpreted as a migration or the change due to increase social instability or worsen health.


### A note on units ##

It is important to understand that:

   **population** is in number of people
   **nature and wealth ** in dollars
   **time** in years
   
We can argue that nature and wealth can be adjusted by prices, and the natural evaluation can be very complex, and its dependency and disccounting is subject to a great deal or intergenerational values.  


### Equilibrium

The Handy model can be used to find sustainable equilibriums and the maximum carrying capacity for a given society.
In order for the population to reach an equilibrium, we must have:


$$ 
\alpha_m \le \beta_E \le \beta_C \le \alpha_M   \   [13]   
$$
   
 We define a dimensionless parameter:
 
 
$$ 
\varphi =(\alpha_M- \beta_C)/ (\alpha_M - \alpha_m)     \   [14]   
$$

Since 

$$ 
\alpha_m  \le \beta_C \le \alpha_M   \     
$$
That means that:

$$ 
0 \le \varphi \le 1      \     
$$

### Equilibrium in a egalitarian society , xe=0

We can find the equilibrium values where [4][6] and [7] are equal to zero, that gives us:

$$
x_{C,e}= (\gamma/\delta)*(\lambda-\varphi*(s/\delta))  \ [15]
$$

$$
y_{e}= \varphi*(s/\delta))  \ [16]
$$
$$
w_{e}= \varphi\rho*x_{Ce}  \ [17]
$$

People carrying capicity just equals Xce

$$
\chi  = x_{C,e}= (\gamma/\delta)*(\lambda-\varphi*(s/\delta))  \ [18]
$$

Carrying capcity can be maximized if nature regeneration rate is maximal, that happens when:


$$
y_{e}= \lambda /2  \
$$

This requires delta to be set at an optimal value equal to the maximum sustainable population, which can be derived from equations 15,16,17.

$$
\delta_{opt}= (2*\varphi*s)/\lambda  \
$$

The maximum carrying capcity is thus:

$$
\chi_M  = (\gamma/\delta_{opt})*(\lambda-\varphi*(s/\delta_{opt})) = (\gamma/(\varphi*s))*(\lambda/2)^2
$$

Following the same procedure, the equilibrium values of population, nature and wealth can be estimated, together with the maximum natural capacity for the following scenarios, proposed in Handy:

  A) Egalitarian society, where Xe=0
  B) Equitable societt, xe>0 but kappa=1
  C) Unequal society, xe>0 and kappa>1
  
For all of these scenarios, the typical parameter values and initial conditions are given as following

```{r,, echo=FALSE,message=FALSE,warning=FALSE}
#install.packages("knitr")

library(knitr)
library(dplyr)
library(kableExtra)
param <-read.csv("data/Parameters_initial_values.csv")



kable(param) %>%
   kable_styling(bootstrap_options = "striped", full_width = F)
```






